---
title: "Data Cleaning"
author: "Ashley, Zoey, Vinithra, Hailey"  
html:  
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  echo: true
  warning: false
  message: false
  error: TRUE
---

## Data Cleaning

**Loading Required Packages**

```{r}
library(tidyverse)
library(readr)
library(knitr)
```

**Life Expectancy Data** 

Here I will load and clean the life expectancy data for the project.I decided to remove all years with missing values, and decrease sample size to around 100 data points. In addition, I also created a separate table of average life expectancy for each country per decade.

```{r}
lex <- read_csv("lex.csv")
lex_long <- lex |>
  select(country,"1950":"2050") |>
  pivot_longer(cols = !country, 
               names_to = "year", 
               values_to = "life_ex")
```




```{r}
avg_lex_decade <- lex |>
  mutate("1950s" = rowMeans(across("1950":"1959"))) |>
  mutate("1960s" = rowMeans(across("1960":"1969"))) |>
  mutate("1970s" = rowMeans(across("1970":"1979"))) |>
  mutate("1980s" = rowMeans(across("1980":"1989"))) |>
  mutate("1990s" = rowMeans(across("1990":"1999"))) |>
  mutate("2000s" = rowMeans(across("2000":"2009"))) |>
  mutate("2010s" = rowMeans(across("2010":"2019"))) |>
  mutate("2020s" = rowMeans(across("2020":"2029"))) |>
  mutate("2030s" = rowMeans(across("2030":"2039"))) |>
  mutate("2040s" = rowMeans(across("2040":"2049"))) |>
  select(1,"1950s":"2040s")
kable(head(avg_lex_decade))
```

**Income Share of Poorest 20%**

Loading and cleaning of data for the income share of the poorest 10% by country.

```{r}
income_share_poorest <- read_csv("income_share_of_poorest_10percent.csv")


```



